---
title: "Downloading UNPD data into R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading-UNPD-data-into-R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`DDSQLtools` is a package that aims to make downloading data from the UNPD SQL API into R easy. In this vignette we will discuss the different ways in which you can explore and download this data. Make sure you have the latest version of `DDSQLtools` by installing it from Github:

```{r, eval = FALSE}
devtools::install_github("timriffe/DDSQLtools")
```

and load the package:

```{r setup}
library(DDSQLtools)
```

## Quick recipes

If you're already familiar with concepts from the UNPD SQL API, then you'll only have to interact with the `get_recorddata` function. For example, here's how to get data on population by age and sex for Egypt with different options:

```{r, eval = FALSE}
egypt_str <- get_recorddata(dataProcessIds = "Census",
                            indicatorTypeIds = "Population by age and sex",
                            locIds = "Egypt",
                            locAreaTypeIds = "Whole area",
                            subGroupIds = "Total or All groups",
                            isComplete = "Abridged")

```

`get_recorddata` accepts either string names or codes. For example, this will return the same thing as above:

```{r, eval = FALSE}
egypt_num <- get_recorddata(dataProcessIds = 2,
                            indicatorTypeIds = 8,
                            locIds = 818, 
                            locAreaTypeIds = 2,
                            subGroupIds = 2,
                            isComplete = 0)
```

Note that **using numeric codes is faster than using string names because we don't have to make code translations behind the scenes** (if you're unsure what these codes mean, please refer to the section `Exploring the UNPD SQL API from R` below). Here are a few other examples:

#### Population by age and sex with census data

```{r}
china <- get_recorddata(dataProcessIds = "Census",
                        indicatorTypeIds = "Population by age and sex",
                        locIds = "China",
                        locAreaTypeIds = "Whole area",
                        subGroupIds = "Total or All groups",
                        isComplete = "Abridged")

head(china)
```

#### Births by age of mother

```{r, error = TRUE}
get_recorddata(dataProcessIds = "Census",
               indicatorTypeIds = "Births by age of mother",
               locIds = "Togo",
               locAreaTypeIds = "Whole area",
               subGroupIds = "Total or All groups",
               isComplete = "Abridged")

```

See the error? This means that using `"Census"` for this set of options is unavailable. We need to check which other data processes have data for these options (for example, using `get_dataprocess()`). Since this is a trial and error process, we've nailed it down to `"Survey: cross-sectional survey (generic/national)"`:

```{r }
togo <- get_recorddata(dataProcessIds = "Survey: cross-sectional survey (generic/national)",
                       indicatorTypeIds = "Births by age of mother",
                       locIds = "Togo",
                       locAreaTypeIds = "Whole area",
                       subGroupIds = "Total or All groups",
                       isComplete = "Abridged")

head(togo)
```

#### Deaths by age and sex using register data

```{r }
oman <- get_recorddata(dataProcessIds = "Register",
                       indicatorTypeIds = "Deaths by age and sex",
                       locIds = "Oman",
                       locAreaTypeIds = "Whole area",
                       subGroupIds = "Total or All groups",
                       isComplete = "Abridged")

head(oman)
```


#### Change the IP of the UNPD SQL API

Currently the IP of the server is automatically set but can be changed simply by adding `options(unpd_server = "your server")` to the beginning of your script:

```{r, eval = FALSE}
# Change this to the desired server. This is the server
# which is set by default.
options(unpd_server = "http://24.239.36.16:9654/un3/api/")

get_recorddata(dataProcessIds = "Register",
               indicatorTypeIds = "Deaths by age and sex",
               locIds = "Oman",
               locAreaTypeIds = "Whole area",
               subGroupIds = "Total or All groups",
               isComplete = "Abridged")

```

#### Downloading many countries

Due to the nature of the UNPD API, it's not possible to download over 30 countries in the same API request. For example, if we wanted to download data on these countries:

```{r, eval = FALSE, error = TRUE}

myLocations <- c(28, 492, 570, 674, 308, 96, 196, 8,
                 376, 662, 670, 642, 84, 188, 442, 100,
                 192, 170, 414, 616, 320, 480, 218,
                 818, 222, 300, 558, 40, 52, 348, 470,
                 620, 702, 858, 56, 780, 388, 246, 352,
                 591, 144, 862)

get_recorddata(dataProcessIds = "Register",
               indicatorTypeIds = "Deaths by age and sex",
               locIds = myLocations,
               locAreaTypeIds = "Whole area",
               subGroupIds = "Total or All groups",
               isComplete = "Abridged",
               startYear = 1920,
               endYear = 2020)

```

You'll get an error saying that `DDSQLtools` could not connect to the API. When that happens you can loop over chunks of countries to get it faster. The first thinkg you want to do is to copy the **optimized query** printed by `get_recorddata` that already translated the string codes to numbers.

##### Using base R

```{r, echo = FALSE, eval = FALSE}
myLocations <- c(28, 492, 570, 674, 308, 96, 196, 8,
                 376, 662, 670, 642, 84, 188, 442, 100,
                 192, 170, 414, 616, 320, 480, 218,
                 818, 222, 300, 558, 40, 52, 348, 470,
                 620, 702, 858, 56, 780, 388, 246, 352,
                 591, 144, 862)

# Here replace with number of desired chunk of countries
n_chunks <- 3
chunk_groups <- rep(1:n_chunks, length.out = length(myLocations))
cnty_groups <- split(myLocations, chunk_groups)

# Loop through each chunk of countries with `lapply`
deaths <- lapply(cnty_groups, function(x) {

  res <- get_recorddata(dataProcessIds = 9,
                        startYear = 1920,
                        endYear = 2020,
                        indicatorTypeIds = 20,
                        isComplete = 0,
                        locIds = x,
                        locAreaTypeIds = 2,
                        subGroupIds = 2)

  # Informative message to keep track of which chunk
  # of countries have been completed.
  cat("Countries", paste0(x, collapse = ", "), "done")
  res
})

# Row-bind all the data frames into one data frame
deaths <- do.call(rbind, deaths)
head(deaths)
```

##### Using the tidyverse
```{r, echo = FALSE, eval = FALSE}
library(tidyverse)

myLocations <- c(28, 492, 570, 674, 308, 96, 196, 8,
                 376, 662, 670, 642, 84, 188, 442, 100,
                 192, 170, 414, 616, 320, 480, 218,
                 818, 222, 300, 558, 40, 52, 348, 470,
                 620, 702, 858, 56, 780, 388, 246, 352,
                 591, 144, 862)

n_chunks <- 3
chunk_groups <- rep(1:n_chunks, length.out = length(myLocations))

cnty_groups <- split(myLocations, chunk_groups)

# Loop through each location with `purrr`
deaths <- map(cnty_groups, ~ {

  res <- get_recorddata(dataProcessIds = 9,
                        startYear = 1920,
                        endYear = 2020,
                        indicatorTypeIds = 20,
                        isComplete = 0,
                        locIds = .x,
                        locAreaTypeIds = 2,
                        subGroupIds = 2)

  cat("Countries", paste0(x, collapse = ", "), " done")
  res
})

deaths <-
  deaths %>%
  bind_rows() %>%
  as_tibble()

deaths
```

#### Identifying ID and Names of values

Whenever you download some data, you'll see that there are several columns that end with `Name`. This columns contain, for example, the name of the location (`LocName`) or the name of the subgroup (`SubGroupNam`). To be able to see the equivalent ID of these names, there are two different strategies using the `haven` package.

First, let's get some data:

```{r }
res <- get_recorddata(dataProcessIds = 9, # Register
                      startYear = 1920,
                      endYear = 2020,
                      indicatorTypeIds = 14, # Births by sex
                      isComplete = 2, # Total
                      locIds = 28, # Antigua and Barbuda
                      locAreaTypeIds = 2, # Whole area
                      subGroupIds = 2) # Total
```

#### Base R

You can use `print_labels` from `haven` on the column of interest and it will show you the ID:

```{r}
library(haven)
print_labels(res$LocName)
```

If you want to flip the labels for the entire data frame such that the ID is current value and not the name, you can do that with `as_factor`:

```{r}
res_edit <- as_factor(res)

# No long showing the name but the ID
head(res_edit["LocName"])
```

#### Tidyverse

Using the tidyverse, the solution from base R applies as well. However, if you convert your data frame to a `tibble`, it will automatically show you the labels in the data frame.

```{r }
library(tidyverse)

res_tb <- res %>% as_tibble()
# See the code next to the name? No need to print it
# separately, `tibble` print boths things together.
res_tb %>%
  select(ends_with("Name"))
```

If you're interested in keeping either the codes or the names, you can do that with `as_factor`:

```{r }
res_tb %>%
  # You can extract just the ID
  as_factor(levels = "labels") %>% 
  select(ends_with("Name"))
```

```{r}
res_tb %>%
  # Or just the values
  as_factor(levels = "values") %>%
  select(ends_with("Name"))  
```

## Exploring the UNPD SQL API from R

The previous examples assume you're familiar with the names or codes of the parameters of the API: locations, subgroups, etc... Whenever you don't know these values, you need to resort to the list of available values. Each parameter defined above has an equivalent `get_*` function to extract all possible values. The current list is:

* dataProcessIds = `get_dataprocess()`
* indicatorTypeIds = `get_indicatortypes()`
* locIds = `get_locations()`
* locAreaTypeIds = `get_locationtypes()`
* subGroupIds = `get_subgroups()`
* isComplete = `"Abridged"`, `"Complete"` or `"Total"`

For example, to see all possible indicator types:

```{r, eval = FALSE}
get_indicatortypes()
```

Here we can see both the **code** and the **name** of the indicator. Both values are accepted by `get_recorddata`. Another example is getting the list of location types. `get_locationtypes()` will give you those:

```{r}
get_locationtypes()
```

In order to search for a specificic combination of values (`"Egypt"`, `"Deaths by age and sex"`, etc...), the process is only possible through trial and error. For example, if we wanted to explore `Deaths by age and sex` for `Oman`, we'd have to explore whether there is data for `Semi-rural` (`locAreaTypeIds`) and `Christian Orthodox` (`subGroupIds`). Since this exploration process is a bit cumbersome, we're currently developing a new strategy for exploring country/indicator pairs that will be more intuitive.

For now, let's try this example for `"Semi-rural"`:

```{r, error = TRUE}
get_recorddata(dataProcessIds = "Register",
               indicatorTypeIds = "Deaths by age and sex",
               locIds = "Oman",
               locAreaTypeIds = "Semi-rural",
               subGroupIds = "Total or All groups",
               isComplete = "Abridged")
```

This error is just saying that the combination of data that you've specified is not available (in this case, `Semi-rural` is not available for register data on deaths by age and sex in `Oman` for all groups abridged). 

Figuring out which specific set of values returns what the user want is a tedious process. If the user doesn't know the specific combination of values by memory, the most optimal strategy that we advise is to explore the available values in the [front end of the API](http://24.239.36.16:9654/un3/Web/index.html) and then use these in the request.

Let's do one example. Follow the hyperlink and click on `"Get Started - Select Indicator & Country"`. Then select `"Deaths by age and sex"` -> `"Oman"`. All possible datasets for `Oman` and `"Deaths by age and sex"` will appear automatically. We can filter down the options in the left-most menu and use these values to populate our `get_recorddata` call.

Looking at the possible values of `"Type of location"`, only `"Whole area"` is available. This means that we can't specify `"Semi-rural"` (as we saw in the example above). Looking at `"Type of subgroup"`, the only available group is `Total or All groups`, meaning that we can't specify `"Christian Orthodox"` either.

Using this strategy is much easier and less time consuming that trying out different values. As outlined above, this strategy will not be necessary in the near future as we develop a quick strategy for exploring data from R.

